Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics

### 2013-08-14 paper/paper2.tex
Added more text, and a new figure describing overfitting

Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high-resolution models that maintain quantitative agreement with the underlying simulation data and low-resolution models which capture the salient features of the potential energy landscape while sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states. The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which determine the kinetics.

@@

Here, we seek to build models that are \emph{suitably} complex, given the data, yielding complex descriptions of the system only to the extent that their additional parameters are implied by the observed dynamics. To that end, we introduce a new procedure for scoring the likelihood of an MSM, which, together with the Bayesian information criterion (BIC), enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states. This approach complements validation procedures performed primarily based on human intuition, such as Chapman-Kolmogorov tests, and enables us to treat model selection as an optimization problem amenable to automated methods.

@@

The appropriate likelihood is a path action in (continuous) phase space, on which the discrete states are merely an indicator function basis.

\begin{align}
P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(X_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^{T} p_{s(x_i)}(X_{i})
\label{eq:like}
\end{align}

@@

However, models' long timescale behavior -- the rates and fluxes between metastable basins are independent of the choice of the emission distributions. The emission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state. If such differences exist and are sufficiently large to warrant attention, than the states are too large.

Therefore, the most appropriate emission distribution for discrete state MSMs is that of the uniform distribution over the phase-space volume of the state. That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state i is 0 if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume. This constant is set so that the distribution integrates to 1, and is thus the reciprocal volume of the microstate.

@@

\section{Model Selection With the Bayes Information Criterion}

Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Hastie01Elements}. Statistical learning theory provides a number of alternative approaches for this problem. Conceptually, the most straightfoward is a full Bayesian treatment in which all unknown model parameters are representated by probability distributions. The evidence for a model is computed by formally integrating over the model parameters. The evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space. Unfortunately such approaches are intractable for problems of this size because of the need to integrate over all possible Markov models of a given size.

Instead, we find that Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} provides an efficient framework for choosing the number of states in a Markov state model.

What  about cross validation?

@@

\begin{figure}
\centering
\vspace{2em}
\includegraphics[width=3in]{figs/fig1_rbow.png}
\caption{Three systems studied in this work. (a) Langevin dynamics on the two dimensional M\"{u}ller potential. (b) 50 ns of dynamics of alanine dipeptide. (c) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{}, courtesy of D.E. Shaw research \cite{}.}
\end{figure}

\section{Results and Discussion}

\bibliography{bibliography}

\end{document}

>>>>

Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models. Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states. The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which determine the kinetics.

@@

Here, we seek to build models that are \emph{suitably} complex, given the data, yielding complex descriptions of the system only to the extent that their additional parameters are implied by the observed dynamics. To that end, we introduce a new procedure for scoring the likelihood of an MSM, which, together with cross validation and the Bayesian information criterion (BIC), enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states. This approach complements validation procedures performed primarily based on human intuition, such as Chapman-Kolmogorov tests, and enables the treatment of model selection as an optimization problem amenable to automated methods.

@@

The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis. With $s(X)$ as the function mapping conformations into the indicator function basis set, $s : \mathbb{R}^{3N} \rightarrow \{1, 2, \ldots, K\}$, the likelihood can be written as

\begin{align}
P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^{T} p_{s(x_i)}(x_{i})
\label{eq:like}
\end{align}

@@

However, models' long timescale behavior -- the rates and fluxes between metastable basins -- are independent of the choice of the emission distributions. The emission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state. If such differences exist and are sufficiently large to warrant attention, than the states are too large.

Therefore, the most appropriate emission distribution for discrete state MSMs is that of the uniform distribution over the phase-space volume of the state. That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state $i$ is $0$ if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume. This constant is set so that the distribution integrates to $1$, and is thus the reciprocal volume of the microstate.

@@

\section{Cross Validation and the Bayes Information Criterion}

\begin{figure}
\centering
\includegraphics[width=3.1in]{figs/overfitting.png}
\caption{An example of overfitting. Left: in the presence of noise, a more complex model is able to fit the observed training data better, with lower residues and a higher empirical likelihood, but fails to distinguish the signal from the noise. Right: more complex model classes exhibit lower training errors, but generalize poorly to unobserved data.}
\end{figure}

Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to overfitting\cite{Hastie01Elements}. Statistical learning theory provides a number of alternative approaches for this problem. Conceptually, the most straightforward is a full Bayesian treatment in which all unknown model parameters are represented by probability distributions. The evidence for a model is computed by formally integrating over the model parameters and the evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space. Unfortunately such approaches are intractable for problems of this size because of the need to integrate over all possible Markov models of a given size.

Instead, we explore both cross validation and Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} for choosing the number of states in a Markov state model. For cross validation, we parameterize the model, building both the state space and transition matrix on a subset of the data, but evaluating the likelihood on the left-out portion in an attempt to directly measure the generalization error. The BIC on the other hand involves augmenting the likelihood with a penalty on the number of parameters, and is
an asymptotic approximation Bayesian evidence.

\begin{equation}
\label{eq:bic}
\mathrm{BIC} \equiv -2\cdot \ln L + k \ln (n)
\end{equation}

where $L$ is the likelihood, $k$ is the number of free parameters, and $n$ is the number of data points, assumed to be independent and identically distributed.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs/fig1_rbow.png}
\caption{Three systems studied in this work. (a) Langevin dynamics on the two dimensional M\"{u}ller potential. (b) 50 ns of dynamics of alanine dipeptide. (c) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{}.}
\end{figure}

The uniform distribution emission model presents a computational challenge: its use requires the calculation of the (hyper)volume of the MSM's states, which, when defined by clustering are high-dimensional Voronoi cells. While trivial in two or three dimensions, this computational geometry task becomes challenging in high-dimensional settings. The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{}. We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics dataset is first projected into a suitable small vector space of up to perhaps ten dimensions.

A further challenge is the procedure by which to model the volume of states which are at the ``edge'' of the MSM -- whose Voronoi cells extend to infinity in some direction. Is the volume of these states unbounded? It is appropriate to assert that the volume of these edge states is bounded in some way by the extent of our dataset. For example, the volume of a state might be defined as the volume of the intersection of its Voronoi cell and the convex hull of the whole dataset, which would encode the assumption that the likelihood of observing conformations outside the convex hull of the sampled data is vanishing.

We use a slightly modified version of this definition that adopts the same spirit. Instead of taking the outer bounding envelope to be the convex hull of the data, we take it to be the set of all trial points such that the nearest sampled configuration to the trial point is closer than a certain cutoff, $R$. For further efficiency, it is feasible to use only a random subsample of the dataset for this nearest neighbor computation.

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated [HOW MANY STEPS?] of Langevin dynamics on the M\"{u}ller potential, clustered the data using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with [HOW MANY ROUNDS] of Monte carlo rejection sampling, using a padding distance of $R=?$ around the state centers from the 500 state model to define the datasets envelope.

\begin{figure}[h]
\centering
\includegraphics[width=2in]{figs/like_comp.png}
\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states, very close to the region in which the BIC is minimized. The BIC optimal model is also where the models' slowest implied timescals begins to become invariant with respect to the number of states.}
\label{fig:mullerlike}
\end{figure}

As shown in Fig. \ref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly. Here, the BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate. This model selection is consistent with tests based on the invariance of transition matrix eigenspectrum, but permits a direct optimization of the state space.

\subsection{Alanine Dipeptide}

\subsection{Fip35 WW Domain}

\begin{figure}
\centering
\includegraphics[width=2in]{figs/ww_bic_vs_eval_slide_nscounts.png}
\caption{}
\label{fig:ww}
\end{figure}

To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw research \cite{}. In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis, retaining between two and six uncorrelated(?) linear combination of pairwise distances(?).

\bibliography{bibliography}
\end{document}
